---
title: "Modeling Workforce Retention in R"
author: "Gordon Goodwin"
format: html
code-fold: true
code-link: true
code-line-numbers: true
theme: cerulean
toc: true
toc-depth: 4
toc-title: "Contents"
toc-location: right
number-sections: false
editor: visual
self-contained: true
whitespace: small
link-external-icon: true
execute: 
  warning: false
  error: false
  cache: false
---

This guide illustrates how **survival analysis** techniques can be used to analyze & predict **employee attrition & retention** using `R`.

# Case Study

For this guide, we'll use an illustrative case study based on real scenarios from my professional experience.

::: callout-note
## The data used here can be found on my [Github page](https://github.com/gordon-goodwin/gordon-goodwin.github.io/tree/main/R_projects/data) and was simulated to protect confidentiality, but still reflects real patterns and insights. Please reach out with Q's!
:::

## Context

GMG Automotives is a regional automotive manufacturer with 3 production facilities in `Denver CO`, `Tampa FL`, and `Charlotte NC`.

-   All 3 facilities have been open \~1 year (opened June 2023).

-   All 3 facilities have a "cohort-style" hiring strategy, with 2 main hire classes that started in June 2023 and Jan 2024 respectively.

-   Plant Mgmt had hoped that the Jan 2024 hire class would be sufficient to maintain sufficient staffing levels for at least 6 months (June 2024).

## Problem

Over last few months, the Chief HR Officer (CHRO) has received anecdotal reports from plant management that [employee turnover is alarmingly high and trending upwards]{style="color:red"}

-   The Talent Aquisition (TA) team is [having difficulty drafting a hiring strategy]{style="color:red"} that will support a [stable average headcount of 200 employees at each facility]{style="color:green"}

## Objective

Model **employee attrition**, identify **risk factors**, and **predict retention probabilities** over time.

### Business Questions

-   [*What % of employees can be expected to retain past the training (30d) and probationary (90d) periods?*]{style="color:blue"}

-   [*Are there any risk factors that may impact the probability an employee retains?*]{style="color:blue"}

    -   Facility (`Denver`, `Tampa`, `Charlotte`)

    -   Hire Cohort (`June 2023`, `Jan 2024`)

-   [*How can we leverage this information to inform workforce planning strategy?*]{style="color:blue"}

    -   Goal = stable headcount of 200 employees at each facility

------------------------------------------------------------------------

## Data

We have obtained basic HR data that reflects the most recent employee information, `employment status`, and `seniority time` available for each employee.

```{r}
#| label: setup

# No Scientific Notation
options(scipen = 999)

# Source Functions
library(tidyverse)
library(tidymodels)
library(censored)
library(janitor)
library(ggfun)
library(survminer)
library(ggsurvfit)
library(patchwork)
library(gt)

# Read Data
read_csv("Survival Analysis PA Tutorial.csv") |> clean_names() -> dat1

# Format
dat1 |> 
  mutate(hire_date = mdy(hire_date),
         max_date_recorded = mdy(max_date_recorded),
         hire_date = factor(hire_date),
         facility = factor(facility),
         transfer = factor(transfer),
         manager = factor(manager),
         termination_status = factor(termination,
                                     labels = c("Active","Termed"))) -> dat1


# Preview
dat1 |> 
  select(emp_id, facility, manager, hire_date,
         seniority_recorded, termination_status) |> 
  gt::gt_preview()

```

This data format is known as **time-to-event** data, as it allows us to answer questions about how long an object or subject can be expected to "survive" before some event of interest occurs.

[In employee attrition contexts, the "subjects" are employees, the "event" is termination, and our objective is to **model how long employees can be expected to retain before terminating**.]{style="color:purple"}

# Results

Our results are organized into 3 main sections:

1.  **Headcount Trends**
2.  **Baseline Retention**
3.  **Group Differences**

## Headcount

***Have staffing thresholds consistently been met in 2024 across all 3 facilities?***

Visualizing our headcount trends across facilities reveals [**2 main problems**]{style="color:red"}:

-   Both the **`Tampa FL`** and **`Charlotte NC`** facilities currently have staffing levels [**below** the 200-employee threshold needed to sustain operations]{style="color:red"}

-   While all 3 facilities have followed the same hiring strategy, differences in attrition have led to [headcount discrepancies that **get wider over time**]{style="color:red"}

```{r}
#| label: headcount

# Read Data
read_csv("Headcount Trend PA Tutorial.csv") |> clean_names() -> dat2

# Format
dat2 |> 
  mutate(date = mdy(date),
         facility = factor(facility)) -> dat2

# Plot
dat2 |> 
  ggplot(aes(x = date,
             y = headcount,
             color = facility,
             group = facility)) +
  geom_point() +
  geom_line(linewidth = 0.75) +
  scale_x_date(date_breaks = "1 month",
               date_labels = "%b-%Y") +
  scale_y_continuous(breaks = seq(0, 300, 50),
                     limits = c(0, 300)) +
  scale_color_manual(values = c("#663171", "#ea7428","#0c7156")) +
  geom_hline(yintercept = 200, lty = 2, linewidth = 0.75, color = "blue") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top",
        plot.title = element_text(face = "italic",color = "firebrick"),
        axis.text.x = element_text(angle = 45, vjust =1, hjust=1)) +
  labs(x = "Date", y = "Headcount", color = "Facility",
       title = "Headcount by Facility")

```

## Baseline Retention

***How long can the typical GMG Auto employee be expected to retain after hire?***

Below, our **baseline retention curve** displays the probability of an employee retaining past a given seniority point:

-   [*Roughly **\~80%** of employees can be expected to **retain past training (30d)***]{style="color:purple"}

-   [*Around **\~60%** of employees can be expected to **retain past probation (90d)***]{style="color:blue"}

-   [*A little over **40%** of employees can be expected to **retain past 6m***]{style="color:orange"}

-   [*Less than **20%** of employees can be expected to **retain past 12m***]{style="color:red"}

```{r}
#| label: baseline retention

# Plot Survival Curve using prebuilt fns
survfit2(Surv(seniority_recorded, termination) ~ 1, data = dat1) |>
  ggsurvfit() +
  labs(
    x = "Seniority Days",
    y = "Retention Probability",
    title = "GMG Automotive Baseline Employee Retention"
  ) +
  add_confidence_interval() + 
  # add_censor_mark() +
  # add_risktable(risktable_stats = c("n.risk",
  #                                  "n.event")) +
  scale_y_continuous(labels = scales::label_percent(),
                     breaks = seq(0, 1, .1),
                     limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 365, 30),
                     limits = c(0, 365)) +
  geom_vline(xintercept = 30, color = "red", linewidth = .8, lty = 2) +
  geom_vline(xintercept = 90, color = "sienna", linewidth = .8, lty = 2) +
      geom_vline(xintercept = 180, color = "purple", linewidth = .8, lty = 2) +
    geom_vline(xintercept = 365, color = "green4", linewidth = .8, lty = 2) +
  theme(plot.title = element_text(face = "italic"))
```

We can also format this retention curve as a **survival table**:

```{r}
#| label: overall survtable

## Fit Kaplan Maier Survival Model
survfit(Surv(time = seniority_recorded,
             event = termination) ~ 1,
        data = dat1) -> survmod_overall

# Table of summary results
data.frame(time = survmod_overall$time,
           n_risk = survmod_overall$n.risk,
           n_event = survmod_overall$n.event,
           n_censored = survmod_overall$n.censor,
           survival = survmod_overall$surv,
           std_error = survmod_overall$std.err,
           ci_95_lower = survmod_overall$lower,
           ci_95_upper = survmod_overall$upper) |> 
  mutate(across(where(is.numeric),
                ~round(.x, 2)))-> survtab_overall

survtab_overall |> 
  filter(time %in% c(90, 180, 365)) |> 
  select(`Seniority Days` = time, 
         `Retention Rate` = survival,
         `Conf Int Lower` = ci_95_lower,
         `Conf Int Upper` = ci_95_upper) |> 
  gt() |> 
    tab_header(title = "Retention at 3m, 6m, & 12m Seniority",
               subtitle = "GMG Automotive Employees") |> 
  fmt_percent(columns = c(`Retention Rate`), decimals = 0)

```

## Group Differences

In order to probe for **group differences** and **risk factors** in attrition and retention rates, we've now calculated **group-level** retention curves and tables**.**

### Facilities

***Do retention rates vary by facility? Can an employee's facility predict their odds of terminating?***

The **retention curve** below supports the following insights:

-   *Retention rates **do** vary significantly by facility*

    -   Employees at the **`Denver, CO`** facility have the **greatest chance of retaining** over time.

    -   Employees at the **`Charlotte NC`** facility have the **lowest chance of retaining** over time.

    -   **`Tampa FL`** falls somewhere in the middle

```{r}
#| label: facilitysurvival


# Plot Curve - ggsurvfit
survfit2(Surv(time = seniority_recorded,
              event = termination) ~ facility,
         data = dat1) |> 
  ggsurvfit(linewidth = 0.8) +
  # add_confidence_interval() + 
  add_pvalue() +
  scale_y_continuous(labels = scales::label_percent(),
                     breaks = seq(0, 1, .1),
                     limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 365, 30),
                     limits = c(0, 365)) +
  geom_vline(xintercept = 30, color = "red", linewidth = .65, lty = 2) +
  geom_vline(xintercept = 90, color = "sienna", linewidth = .65, lty = 2) +
      geom_vline(xintercept = 180, color = "purple", linewidth = .65, lty = 2) +
    geom_vline(xintercept = 365, color = "green4", linewidth = .65, lty = 2) +
  scale_color_manual(values = c("#663171", "#ea7428","#0c7156")) +
  scale_fill_manual(values = c("#663171", "#ea7428","#0c7156")) +
  labs(x = "Seniority Days",y = "Retention Probability",
       color = "Facility",fill = "Facility",
       title = "Employee Retention by Facility") +
  theme(legend.position = "top",
        plot.title = element_text(face = "italic"),
        legend.key = element_rect(fill = "grey90"))

```

As seen below, we can also use a **hazard model** to quantify the impact that an employee's facility assignment has upon their **odds of terminating**.

Using `Denver CO` as the reference group, we can see that:

-   *Employees at the* **`Tampa FL`** *facility are **\~1.6x more likely to terminate** over time than their* **`Denver CO`** *peers.*

-   *Employees at the* **`Charlotte NC`** *facility are **\~2.4x more likely to terminate** over time than their peers at the* **`Denver CO`** facility.

```{r}
#| label: facility hazards regression

# Facility as a covariate for the hazard of termination
coxmod_facility <- coxph(Surv(time = seniority_recorded,
                               event = termination) ~ facility,
                          data = dat1)

# Output
coxmod_facility |> gtsummary::tbl_regression(exp = T) 

```

### Hire Cohorts

***Is employee retention getting worse? Are the Jan 2024 hires more likely to terminate?***

The **retention curve** below supports the following insights:

-   *Employees hired in the* **`Jan 2024`** *cohort **retain worse** than those in the* **`June 2023`** *cohort*

-   *The gap in retention probabilities appears to get **wider over time***

    -   At the **30d** mark, the retention rate for the `June 2023` is **only 7pp greater** than that of the `Jan 2024` cohort (82% vs 75%)

    -   At the **90d** mark, the `June 2023` cohort has a retention rate that is **over 20pp greater** than the `Jan 2024` cohort (70% vs 50%)

```{r}
#| label: cohort survival


# Plot Curve - ggsurvfit
survfit2(Surv(time = seniority_recorded,
              event = termination) ~ hire_date,
         data = dat1) |> 
  ggsurvfit(linewidth = 0.8) +
  # add_confidence_interval() + 
  add_pvalue() +
  scale_y_continuous(labels = scales::label_percent(),
                     breaks = seq(0, 1, .1),
                     limits = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 365, 30),
                     limits = c(0, 365)) +
  geom_vline(xintercept = 30, color = "red", linewidth = .65, lty = 2) +
  geom_vline(xintercept = 90, color = "sienna", linewidth = .65, lty = 2) +
      geom_vline(xintercept = 180, color = "purple", linewidth = .65, lty = 2) +
    geom_vline(xintercept = 365, color = "green4", linewidth = .65, lty = 2) +
  scale_color_manual(values = c("blue", "red","#0c7156")) +
  scale_fill_manual(values = c("blue", "red","#0c7156")) +
  labs(x = "Seniority Days",y = "Retention Probability",
       color = "Hire Class",fill = "Hire Class",
       title = "Employee Retention by Hire Class") +
  theme(legend.position = "top",
        plot.title = element_text(face = "italic"),
        legend.key = element_rect(fill = "grey90"))
```

Similarly, as shown below, a **hazard model** shows that:

-   *Employees from the* `Jan 2024` hire class *are **over 2.4x more likely to terminate** over time than employees hired in the* **`June 2023`** cohort.

```{r}
#| label: cohort hazard

# Hire Cohort as a covariate for the hazard of termination
coxmod_hiredate <- coxph(Surv(time = seniority_recorded,
                               event = termination) ~ hire_date,
                          data = dat1)

# Output
coxmod_hiredate |> gtsummary::tbl_regression(exp = T) 

```

# Conclusions

Our analysis supports 3 primary conclusions & recommendations:

1.  **Attrition & retention rates vary significantly between facilities**, which in turn leads to inconsistent headcount fluctuations.
    1.  [*We should consider interviewing both staff and management at the `Denver CO` facility to understand why their employees retain so much better.*]{style="color:green"}
    2.  [*We should similarly interview the `Charlotte NC` facility to understand why their employees retain so poorly*]{style="color:green"}
2.  **A uniform hiring strategy for all 3 locations will not be sufficient** to maintain a 200-employee threshold at each facility until attrition rates are consistent.
    1.  [*We should consider a 3-tiered approach to hire targets that prioritizes allocation in order of* `Charlotte NC`*,* `Tampa FL`, and then `Denver CO`.]{style="color:green"}
3.  **Underlying attrition rates seem to be getting worse**, as the `Jan 2024` cohort is retaining at a significantly lower rate than the `June 2023` class.
    1.  [*We should consult plant management to see if there has been any external change or seasonal variations in workload that might explain the recent decline*]{style="color:green"}
    2.  [*We should consider increasing the next hire class size above for all 3 facilities to be safe*]{style="color:green"}
